Supplementary material for: non-coding RNAs of bird genomes

### 2014-04-15 paper/manuscript.tex
Responding to referee comments.

%\title{Non-coding RNAs of bird genomes}
%\title{Wins and losses: Non-coding RNAs of bird genomes}
\title{Fly away Peter, come back Paul: the conservation and losses of avian non-coding RNAs}

@@

%% Non-coding RNAs in bird genomes

%% Paul P. Gardner1,2, Sarah W. Burge3, Jana Hertel5, Maria Ninova4, Stephanie Kehr5, Anne Nitsche5, Mario Fasold5, Tammy E. Steeves1, Sam Griffiths-Jones4, Peter Stadler5

%% 1 School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand. 2 Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand. 3 European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK. 4Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom. 5 Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, Hartelstrasse 16-18, D-04107 Leipzig, Germany

%% Talking points:
%% 1. Although the majority of classical eukaryotic ncRNAs are conserved in birds, several notable classical ncRNAs are not. Homology searches in associated genes indicate these apparent "losses" are most likely due to divergence.
%% 2. Of the 59 snoRNA conserved between humans and yeast, we find 45 are conserved in birds. Homology searches in associated genes for the remaining non-conserved snoRNAs indicate these apparent losses are indeed due to multiple losses in several bird lineages.
%% 3. We find that 3 families of broadly conserved vertebrate microRNA families are lost in the avian lineage. We also annotate a number of bird-specific microRNAs.
%% 4. Several well-characterised lncRNA families of known function are conserved between mammals and birds.
%% 5. We discuss the conservation of HOX associated lncRNAs and several lncRNAs that have been implicated in processes associated with cancer.

%% %% Draft Abstract
%% We present the results of a large-scale bioinformatic annotation of non-coding RNAs in 48 avian genomes. We use probabilistic models of hand-curated families from the Rfam database to infer conserved RNA families within each avian genome. We supplement these annotations with predictions from the tRNA annotation tool, tRNAscan-SE, microRNAs from miRBase and selected sequences from analyses of the chicken transcriptome. Although we show extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds, we also demonstrate apparent ``losses'' in several RNA families. These include the divergence of some classical ncRNAs and the genuine loss of several snoRNAs and microRNAs. In contrast, we show a signicant number of lncRNAs are surprisingly well conserved between birds and mammals including several intriguing cases where the reported mammalian lncRNA function is not conserved in birds. These combined results illustrate the utility of applying homology based methods for annotating vertebrate genomes and illustrate many complex evolutionary patterns within the avian ncRNA cohort.

%% LNCRNA:

@@

%%{\bf REDUCE? too much talking about what it can't to compared to what it can do...}
The CM based approach is appropriate for most classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. 
%Recent technological advances have led to
%dramatic speed and memory-usage enhancements for CM analyses
%\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. 
CMs cannot model the exon-intron structures of spliced lncRNAs, nor do
they deal elegantly with the repeats that many lncRNAs
host. Consequently in the latest release of Rfam the lncRNA families
that were added were composed of local conserved (and possibly
structured elements) within lncRNAs, analogous to the ``domains''
housed within protein sequences \cite{Burge:2013}. Whilst some these
regions may not reflect functional RNA elements but instead regulatory
regions, enhancers or insulators, their syntenic conservation still
provides an indication of lncRNA conservation
\cite{diederichs2014four}.

@@

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. 
{\bf Using accurate homology
  search tools and evolutionary constraints, we produce an accurate
  set of genome annotations for 48 predominantly non-model bird
  species for ncRNAs that are conserved across the avian species. This
  conservative set of annotations is expected to contain the core
  avian ncRNAs.}  We focus our report on the unusual results within
the avian lineages. These are either unexpectedly well-conserved
ncRNAs or unexpectedly poorly-conserved ncRNAs. The former are ncRNAs
that are not expected to be conserved between the birds and the other
vertebrates, particularly those ncRNAs whose function is not conserved
in birds.  The latter are apparent losses of ncRNAs expected to be
conserved; Here, we consider three categories of such ``loss'': First,
genuine gene losses in the avian lineage where ncRNAs well conserved
in other vertebrates are completely absent in birds.  Second,
``divergence'' where ncRNAs have undergone such significant sequence
and structural alternations that homology search tools can no longer
detect a relationship between other vertebrate exemplars and avian
varieties.  Third, ``missing'' ncRNAs that failed to be captured in
the available, largely fragmented, avian genomes.  We postulate that
the latter category is likely to be prevalent in comparative avian
genome studies given the distinctive organisation of the avian
genome. Namely, the avian karyotype is characterised by a large number
of chromosomes (average $2n \approx 80$) generally consisting of a few
larger ``macrochromosomes'' and many smaller ``microchromosomes''
\cite{Griffin:2007}. This ’so many, so small’ pattern presents
significant assembly challenges \cite{Ellegren:2005}. Indeed, of the
48 published avian genomes, 20 of which are high-coverage ($>50X$ ),
only two are chromosomally assembled (chicken and zebra finch;
\cite{birds:14,Warren:2010}).

@@

@@

The bulk of the ``unusually well conserved RNAs'' belong to the long
non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
RNAs that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have largely been
characterised in mammalian species, particularly human and
mouse. Consequently, we generally do not expect these to be conserved
outside of mammalia. Notable examples include Xist \cite{Duret:2006}
and H19 \cite{Smits:2008}.  There is emerging evidence for the
conservation of ``mammalian'' lncRNAs in other vertebrates
\cite{Chodroff:2010,Ulitsky:2011}), however, like most lncRNAs, the
function of these lncRNAs remains largely unknown. Here, we show the
conservation of several well-characterised lncRNAs of known function
in humans.

In general, Rfam cannot include the entire length of any large,
spliced RNAs. This is a limitation of the covariance-models used for
the homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently,
only short, well-conserved regions with evolutionarily conserved
secondary structures are included in Rfam. By analogy to
protein-domains, we refer to these as RNA-domains
\cite{Burge:2013}. {\bf SOME OF THESE MAY BE DNA ELEMENTS RATHER THAN
  LNCRNA ELEMENTS!!!!}

@@

The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
\cite{Xu:1997}. Both are expressed in a broad ranch of
tissues. Extensive research on BRCA1 has shown that it is involved in
DNA repair \cite{Moynahan:1999}. The function of NBR2 remains unknown,
yet its conservation across the vertebrates certainly implies a
function (See Figure~\ref{fig:1}).

Of the other classes of RNAs, none showed an unexpected degree of
conservation or expansion within the avian lineage. The only exception
being the snoRNA SNORD93 which has 92 copies in the tinamou
genome, whereas it only has 1-2 copies in all the other vertebrate
genomes.

@@

The overall reduction in avian genomic size has been extensively
discussed elsewhere \cite{Organ:2007}. Unsurprisingly, this reduction
is reflected in the copy-number of ncRNA genes. Some of the most
dramatic examples are the transfer RNAs and pseudogenes which average
$\sim900$ and $\sim580$ copies in the human, turtle and alligator
genomes, the average copies numbers of these drop to $\sim280$ and
$\sim100$ copies in the avian genomes. In addition to reduction in
copy-number, the complete absence of several otherwise ubiquitous
vertebrate ncRNAs in the avian lineage are indicative of genuine gene
loss.

@@

In order to determine to what extent the absence of these RNAs from
the infernal-based annotation is caused by sequence divergence beyond
the thresholds of the Rfam CMs, we complemented our analysis by
dedicated searches for a few of these RNA groups. Our ability to find
additional homologs for several RNA families that fill gaps in the
abundance matrices (Figure ?) strongly suggests that conspicuous
absences, in particular of LUCA and LECA RNAs, are caused by
incomplete data in the current assemblies and sequence divergence
rather then genuine losses.

@@

{\bf Briefly mention the reduction in number of pseudogenes. Pick a
  few key human ones and compare with the birds. }

@@

The ncRNAs presented here have been identified using homology models
and are evolutionarily conserved in multiple avian species. In order
to further validate these predictions we have used strand-specific
total RNA-seq and small RNA-seq of multiple chicken tissues. After
mapping the RNA-seq data to the chicken genome (see Methods for
details), we identified a threshold for calling a gene as
expressed by limiting our estimated false-positive rate to less than
10\%. This FDR was estimated using a negative control of randomly
selected, un-annotated regions of the genome. Since some regions may
be genuinely expressed, the true FDR is potentially much lower than
10\%. Overall, the number of ncRNAs we
have identified in this work that are expressed above background
levels is 865 (72.4\%) (see Table~\ref{table:1}). This shows that 7.0
times more of our ncRNAs are expressed than expected by chance
(Fisher's exact test: $P<10^{16}$). This number is an underestimate of
the fraction of our annotations that are genuinely expressed, as only
a fraction of the developmental stages and tissues of chicken have
been characterized with RNA-seq. Furthermore, some ncRNAs
are expressed in highly specific conditions  \cite{mercer2008specific,johnston2003microrna}.

@@

In this work we have provided a comprehensive annotation of non-coding
RNAs in genome sequences using homology-based methods. The
homology-based tools have distinct advantages over experimental-based
approaches as not all RNAs are expressed in any particular tissue-type
or developmental-stage, in fact some RNAs have extremely specific
expression profiles \cite{Johnston:2003}.  We have identified
previously unrecognised conservation of ncRNAs in avian genomes as
well as some surprising ``losses'' of otherwise well conserved
ncRNAs. We suspect many of these losses are due to a combination of
limitations in the homology search tools that we use for annotation
and the ability of ncRNAs to tolerate large amounts of sequence
variation while remaining functional, rather than \emph{bona fide}
gene loss. In some cases these losses could be due to missing data
from the genome assemblies, but this unlikely to be the case for
multiple independant assemblies.

The collection of ncRNA sequences is generally biased towards model
organisms \cite{Gardner:2010,Hoeppner:2012}. However, we have shown
that using data from well studied lineages such as mammals can result
in quality annotations of sister taxa such as aves.

These results indicate we are still in the very early phases of
determining the functions of many RNA families. This is illustrated by
the fact that the reported functions of some ncRNAs are
mammal-specific, yet are also found in bird genomes.

@@

Bird genomes were searched using the cmsearch program from INFERNAL
1.1 and the covariance models from the Rfam database
v11.0 \cite{Gardner:2011a,Burge:2013}. All matches above the curated GA
threshold were included. Subsequently, all hits with an E-value
greater than 0.0005 were discarded, so only matches which passed the
model-specific GA threshold and had an E-value smaller than 0.0005
were retained. The Rfam database classifies non-coding RNAs into
hierarchial groupings. The basic units are ``families'' which are
groups of homologous, alignable sequences; ``clans'' which are groups
of un-alignable (or functionally distinct), homologous families; and
``classes'' which are groups of clans and families with related
biological functions e.g. spliceosomal RNAs, miRNAs and snoRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013};
these categories have been used to classify our results.

In order to obtain good annotations of tRNA genes we also ran the
specialist tRNA-scan version 1.3.1 annotation tool. This method also
uses covariance models to identify tRNAs. However it also uses some
heuristics to increase the search-speed, annotates the Isoacceptor
Type of each prediction and uses sequence analysis to infer if
predictions are likely to be functional or tRNA-derived pseudogenes
\cite{Lowe:1997,Chan:2009}.

Rfam matches and the tRNA-scan results for families belonging to the
same clan were then ``competed'' so that only the best match was
retained for any genomic region \cite{Gardner:2011a}.  To further
increase the specificity of our annotations we filtered out families
that were indentified in four or fewer of the 51 vertebrate species we
have analysed in this work. These filtered families largely
corresponded to bacterial contamination within the genomic sequences.

999 microRNA sequence families, previously annotated in at least one
vertebrate, were retrieved from miRBase (v19). Individual sequences or
multiple sequence alignments were used to build covariance models with
INFERNAL (v1.1rc3), and these models were searched against the 48 bird
genomes, and the genomes of the american alligator and the green
turtle as outgroups. Hits with e-value <10 realigned with the query
sequences and the resultant multiple sequence alignments manually
inspected and edited using RALEE.

An additonal snoRNA homology search was performed with snoStrip
\cite{Bartschat:2013}. As initial queries we used deutorostomian
snoRNA families from human \cite{Lestrade:2006}, platypus
\cite{Schmitz:2008}, and chicken \cite{Shao:2009}.

@@

The diverse sets of genome annotations were combined. We collapsed the
overlapping annotations into a single annotation. We also generated
heatmaps for different groups of ncRNA genes (see Figure~\ref{fig:1}
and Sup. Figs. 1-3). All the scripts and annotations presented here
are available from Github \cite{gitrepo}.

Chicken ncRNA predictions were validated using two separate RNA-seq
data sets. The first data set (Bioproject PRJNA204941) contains 971
million reads and comprises 27 samples from 14 diferent chicken
tissues sequenced on Illumina HiSeq2000 using a small RNA-seq
protocol. The second data set (personal commun. Matt Schwartz and Igor
Ulitsky) contains 1,46 billion Illumina HiSeq reads sequenced from
whole chicken embryo RNA from 7 stages using a strand-specific dUTP
protocol. Raw reads were checked for quality and adapters clipped if
required by the protocol. Preprocessed reads were mapped to the
galGal4 reference genome using SEGEMEHL short read aligner
\cite{segemehl} and then overlapped with the ncRNA annotations.

@@

Heatmaps showing the prescence/abscence and approximate genomic copy
number of ``unusually, well conserved RNAs'' (particularly the
lncRNAs) on the left and families that have been identified as
surprising RNA Losses, divergence or missing data. In several cases
functionally related families have also been included, e.g. the RNA
components of the major and minor spliceosomes: U1, U2, U4, U5 and U6;
and U11, U12, U4atac, U5 and U6atac, respectively.\label{fig:1}

>>>>

%\title{Non-coding RNAs of bird genomes}
%\title{Wins and losses: Non-coding RNAs of bird genomes}
\title{Conservation and losses of avian non-coding RNAs}

@@

@@

@@

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail.  Using homology search tools
and evolutionary constraints, we produce an accurate set of genome
annotations for 48 predominantly non-model bird species for ncRNAs
that are conserved across the avian species. This conservative set of
annotations is expected to contain the core avian ncRNAs. We focus our
report on the unusual results within the avian lineages. These are
either unexpectedly well-conserved ncRNAs or unexpectedly
poorly-conserved ncRNAs. The former are ncRNAs that were not expected
to be conserved between the birds and the other vertebrates,
particularly those ncRNAs whose function is not conserved in birds.
The latter are apparent losses of ncRNAs expected to be conserved;
Here, we consider three categories of such ``loss'': First, genuine
gene losses in the avian lineage where ncRNAs well conserved in other
vertebrates are completely absent in birds.  Second, ``divergence''
where ncRNAs have undergone such significant sequence and structural
alternations that homology search tools can no longer detect a
relationship between other vertebrate exemplars and avian varieties.
Third, ``missing'' ncRNAs that failed to be captured in the available,
largely fragmented, avian genomes.  We postulate that the latter
category is likely to be prevalent in comparative avian genome studies
given the distinctive organisation of the avian genome. Namely, the
avian karyotype is characterised by a large number of chromosomes
(average $2n \approx 80$) generally consisting of a few larger
``macrochromosomes'' and many smaller ``microchromosomes''
\cite{Griffin:2007}. This ’so many, so small’ pattern presents
significant assembly challenges \cite{Ellegren:2005}. Indeed, of the
48 published avian genomes, 20 of which are high-coverage ($>50X$ ),
only two are chromosomally assembled (chicken and zebra finch;
\cite{birds:14,Warren:2010}).

@@

There is substantial gain and loss of lncRNA
\cite{Cabili:2011,Kutter:2012} and more generally ncRNA loci over
evolutionary time \cite{Hoeppner:2012}. It is difficult to assess how
much of this is due to bioinformatic sequence alignment tools failing
to align correctly below 60-50\% sequence identity
\cite{Gardner:2005:Nucleic-Acids-Res:15860779} or due to genuine gains
and losses. Nevertheless, sequence conservation, generally speaking,
provides useful evidence for gene and function conservation.

We have identified 66,879 loci in 48 avian genomes that share sequence
similarity with previously characterised ncRNAs and are conserved in
$>10\%$ of these avian genomes. These loci have been classified into
626 different families, the majority of which are miRNAs and snoRNAs
(summarised in Table~\ref{table:1}). Out of necessity we have selected
the most charismatic families for further discussion. These are
lncRNAs that appear to be conserved between mammals and aves and the
cases of apparent loss of genes that conserved in most other
vertebrates.

@@

The bulk of the ``unusually well conserved RNAs'' belong to the long
non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
RNAs that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009,Ulitsky:2013}. These RNAs have largely been
characterised in mammalian species, particularly human and
mouse. Consequently, we generally do not expect these to be conserved
outside of mammalia. Notable examples include Xist \cite{Duret:2006}
and H19 \cite{Smits:2008}.  There is emerging evidence for the
conservation of ``mammalian'' lncRNAs in other vertebrates
\cite{Chodroff:2010,Ulitsky:2011}), however, like most lncRNAs, the
function of these lncRNAs remains largely unknown. Here, we show the
conservation of several well-characterised lncRNAs of known function
in humans.

The CM based approach is appropriate for most classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. 
CMs cannot model the exon-intron structures of spliced lncRNAs, nor do
they deal elegantly with the repeats that many lncRNAs
host. Consequently in the latest release of Rfam the lncRNA families
that were added were composed of local conserved (and possibly
structured elements) within lncRNAs, analogous to the ``domains''
housed within protein sequences \cite{Burge:2013}. Whilst some these
regions may not reflect functional RNA elements but instead regulatory
regions, enhancers or insulators, their syntenic conservation still
provides an indication of lncRNA conservation
\cite{diederichs2014four}.

@@

The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
\cite{Xu:1997}. Both are expressed in a broad range of
tissues. Extensive research on BRCA1 has shown that it is involved in
DNA repair \cite{Moynahan:1999}. The function of NBR2 remains unknown,
yet its conservation across the vertebrates certainly implies a
function (See Figure~\ref{fig:1}).

Of the other classes of RNAs, none showed an unexpected degree of
conservation or expansion within the avian lineage. The only exception
being the snoRNA, SNORD93 which has 92 copies in the tinamou
genome, whereas it only has 1-2 copies in all the other vertebrate
genomes.

@@

The overall reduction in avian genomic size has been extensively
discussed elsewhere \cite{Organ:2007}. Unsurprisingly, this reduction
is reflected in the copy-number of ncRNA genes. Some of the most
dramatic examples are the transfer RNAs and pseudogenes which average
$\sim900$ and $\sim580$ copies in the human, turtle and alligator
genomes, the average copy numbers of these drop to $\sim280$ and
$\sim100$ copies in the avian genomes. In addition to reduction in
copy-number, the complete absence of several otherwise ubiquitous
vertebrate ncRNAs in the avian lineage are indicative of genuine gene
loss.

@@

In order to determine to what extent the absence of these RNAs from
the infernal-based annotation is caused by sequence divergence beyond
the thresholds of the Rfam CMs, we complemented our analysis by
dedicated searches for a few of these RNA groups. Our ability to find
additional homologs for several RNA families that fill gaps in the
abundance matrices (Figure~1) strongly suggests that conspicuous
absences, in particular of LUCA and LECA RNAs, are caused by
incomplete data in the current assemblies and sequence divergence
rather then genuine losses.

@@

%{\bf Briefly mention the reduction in number of pseudogenes. Pick a
%  few key human ones and compare with the birds. }

Non-coding RNA derived pseudogenes are a major problem for many ncRNA
annotation projects. The human genome, for example, contains $>1$
million Alu repeats, which are derived from the SRP RNA
\cite{Wheeler:2013}. The existing Rfam annotation of the human genome,
in particular, contains a number of problematic families that appear
to have been excessively pseudogenised. The U6 snRNA, SRP RNA and Y
RNA families have 1,371, 941 and 892 annotations in the human
genome. These are a heterogenous mix of pseudogenised, paralogous,
diverged or functional copies of these families. Unfortunately, a
generalised model of RNA pseudogenes has not been incorporated into
the main covariance model package, Infernal. An approach used by tRNAscan
\cite{Lowe:1997}, is, in theory, generalisable to other RNA families
but this remains a work in progress.

Is it possible that the avian annotations also contains excessive
pseudogenes. However, it has previously been noted that avian genomes
are significantly smaller than other vertebrate species
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}. We
have also noted a corresponding reduction in the number of paralogs
and presumed ncRNA-derived pseudogenes in the avian genomes (See
Supplemental Figure~12). The problematic human families, U6 snRNA, SRP
RNA and Y RNA have, for example, just 26, 4 and 3 annotations
respectively in the chicken genome and 13, 3 and 3 annotations
respectively, on average, in the 48 avian genomes used here.
Therefore, we suspect that the majority of our annotations are in fact
functional orthologs.

@@

The ncRNAs presented here have been identified using homology models
and are evolutionarily conserved in multiple avian species. In order
to further validate these predictions we have used strand-specific
total RNA-seq and small RNA-seq of multiple chicken tissues. After
mapping the RNA-seq data to the chicken genome (see Methods for
details), we identified a threshold for calling a gene as expressed by
limiting our estimated false-positive rate to approximately 10\%. This
FDR was estimated using a negative control of randomly selected,
un-annotated regions of the genome. Since some regions may be
genuinely expressed, the true FDR is potentially lower than
10\%. Overall, the number of ncRNAs we have identified in this work
that are expressed above background levels is 865 (72.4\%) (see
Table~\ref{table:1}). This shows that 7.0 times more of our ncRNAs are
expressed than expected by chance (Fisher's exact test:
$P<10^{16}$). This number is an underestimate of the fraction of our
annotations that are genuinely expressed, as only a fraction of the
developmental stages and tissues of chicken have been characterized
with RNA-seq. Furthermore, some ncRNAs are expressed in highly
specific conditions \cite{mercer2008specific,johnston2003microrna}.

@@

In this work we have provided a comprehensive annotation of non-coding
RNAs in genome sequences using homology-based methods. The
homology-based tools have distinct advantages over experimental-based
approaches as not all RNAs are expressed in any particular tissue-type
or developmental-stage, in fact some RNAs have extremely specific
expression profiles, e.g. the lsy-6 microRNA \cite{Johnston:2003}.  We
have identified previously unrecognised conservation of ncRNAs in
avian genomes and some surprising ``losses'' of otherwise well
conserved ncRNAs. We have shown that most of these losses are due to
difficulties assembling avian microchromosomes rather than \emph{bona
  fide} gene loss. A large fraction of our annotations have been
confirmed using RNA-seq data, which also showed a 7-fold enriched of
expression within our annotations relative to unannotated regions.

The collection of ncRNA sequences is generally biased towards model
organisms \cite{Gardner:2010,Hoeppner:2012}. However, we have shown
that using data from well studied lineages such as mammals can also
result in quality annotations of sister taxa such as aves.

In summary, these results indicate we are in the very early phases of
determining the functions of many RNA families. This is illustrated by
the fact that the reported functions of some ncRNAs are
mammal-specific, yet are also found in bird genomes.

@@

Bird genomes were searched using the cmsearch program from INFERNAL
1.1 and the covariance models from the Rfam database
v11.0 \cite{Gardner:2011a,Burge:2013}. All matches above the curated GA
threshold were included. Subsequently, all hits with an E-value
greater than 0.0005 were discarded, so only matches which passed the
model-specific GA threshold and had an E-value smaller than 0.0005
were retained. The Rfam database classifies non-coding RNAs into
hierarchical groupings. The basic units are ``families'' which are
groups of homologous, alignable sequences; ``clans'' which are groups
of un-alignable (or functionally distinct), homologous families; and
``classes'' which are groups of clans and families with related
biological functions e.g. spliceosomal RNAs, miRNAs and snoRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013};
these categories have been used to classify our results.

In order to obtain good annotations of tRNA genes we ran the
specialist tRNA-scan version 1.3.1 annotation tool. This method also
uses covariance models to identify tRNAs. However it also uses some
heuristics to increase the search-speed, annotates the Isoacceptor
Type of each prediction and uses sequence analysis to infer if
predictions are likely to be functional or tRNA-derived pseudogenes
\cite{Lowe:1997,Chan:2009}.

Rfam matches and the tRNA-scan results for families belonging to the
same clan were then ``competed'' so that only the best match was
retained for any genomic region \cite{Gardner:2011a}.  To further
increase the specificity of our annotations we filtered out families
that were identified in four or fewer of the 51 vertebrate species we
have analysed in this work. These filtered families largely
corresponded to bacterial contamination within the genomic sequences.

999 microRNA sequence families, previously annotated in at least one
vertebrate, were retrieved from miRBase (v19). Individual sequences or
multiple sequence alignments were used to build covariance models with
INFERNAL (v1.1rc3), and these models were searched against the 48 bird
genomes, and the genomes of the American alligator and the green
turtle as outgroups. Hits with e-value $<10$ realigned with the query
sequences and the resultant multiple sequence alignments manually
inspected and edited using RALEE.

An additional snoRNA homology search was performed with snoStrip
\cite{Bartschat:2013}. As initial queries we used deutorostomian
snoRNA families from human \cite{Lestrade:2006}, platypus
\cite{Schmitz:2008}, and chicken \cite{Shao:2009}.

@@

The diverse sets of genome annotations were combined and filtered,
ensuring conservation in 10\% or more of the avian genomes. We
collapsed the remaining overlapping annotations into a single
annotation. We also generated heatmaps for different groups of ncRNA
genes (see Figure~\ref{fig:1} and Supplemental Figure~1-3). All the
scripts and annotations presented here are available from Github
\cite{gitrepo}.

Chicken ncRNA predictions were validated using two separate RNA-seq
data sets. The first data set (Bioproject PRJNA204941) contains 971
million reads and comprises 27 samples from 14 different chicken
tissues sequenced on Illumina HiSeq2000 using a small RNA-seq
protocol. The second data set (personal commun. Matt Schwartz and Igor
Ulitsky) contains 1,46 billion Illumina HiSeq reads sequenced from
whole chicken embryo RNA from 7 stages using a strand-specific dUTP
protocol. Raw reads were checked for quality and adapters clipped if
required by the protocol. Preprocessed reads were mapped to the
galGal4 reference genome using SEGEMEHL short read aligner
\cite{segemehl} and then overlapped with the ncRNA annotations.

@@

Heatmaps showing the presence/absence and approximate genomic copy
number of ``unusually, well conserved RNAs'' (particularly the
lncRNAs) on the left and families that have been identified as
surprising RNA Losses, divergence or missing data. In several cases
functionally related families have also been included, e.g. the RNA
components of the major and minor spliceosomes: U1, U2, U4, U5 and U6;
and U11, U12, U4atac, U5 and U6atac, respectively.\label{fig:1}

### 2014-04-15 paper/supplementary-material.tex
Responding to referee comments.

There are 66 ribosomal modification sites, guided by 59 snoRNA
families, that are preserved between \emph{H. sapiens} and
\emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
families are conserved in the bird dataset.
Over a third of the apparent losses of the yeast-human conserved
snoRNA families appear to cluster on 2 locii of the ancestral
vertebrate genome. We investigated these losses further.

@@

The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned
for specificity and thus misses several occurrances that are easily
found by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
sequences appear degraded at the ends, which may be explained e.g.\ by
low sequence quality at the very ends of contigs or scaffolds. A
\texttt{blastn} search also readily retrieves additional RNase P and
RNAse MRP RNAs, capturing only the best conserved regions. In many
cases these additional candidates are incomplete or contain
undetermined sequence, explaining why they are missed by the
CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
the majority of the genomes. An additional candidate could also be
retrieved for telomerase RNA. Telomerase is well known to exhibit very
poor sequence conservation and rapid variations in size that make it
notoriously hard to identify by homology search \cite{Xie:08a}. The
poor return thus does not come as a surprise. Since \texttt{blastn}
searches remained unsuccessfull we constructed a sauropsid-specific CM
for the vault RNA. In addition to the hits identified by the Rfam
model we obtained three additional homologs. Vault RNAs, with a size
of about 100 nt, exhibit conserved sequence patterns only at their
ends, with essentially unconstrained sequence in the central
part. Their identification is one of the well-know difficult problems
for homology search \cite{Stadler:09b,Kolbe:2009}.

@@

>>>>

There are 66 ribosomal modification sites, guided by 59 snoRNA
families, that are preserved between \emph{H. sapiens} and
\emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
families are conserved in the bird dataset.
Over a third of the apparent losses of the yeast-human conserved
snoRNA families appear to cluster on 2 loci of the ancestral
vertebrate genome. We investigated these losses further.

@@

The simplest case are the selenocysteine tRNAs. Here, tRNAscan is
tuned for specificity and thus misses several occurrances that are
easily found by \texttt{blastn} with $E\le 10^{-30}$. In some cases
the sequences appear degraded at the ends, which may be explained
e.g.\ by low sequence quality at the very ends of contigs or
scaffolds. A \texttt{blastn} search also readily retrieves additional
RNase P and RNAse MRP RNAs, capturing only the best conserved
regions. In many cases these additional candidates are incomplete or
contain undetermined sequence, explaining why they are missed by the
CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
the majority of the genomes. An additional candidate could also be
retrieved for telomerase RNA. Telomerase is well known to exhibit very
poor sequence conservation and rapid variations in size that make it
notoriously hard to identify by homology search \cite{Xie:08a}. The
poor return thus does not come as a surprise. Vault RNA homology
searches with \texttt{blastn} remained unsuccessful, therefore we
constructed a sauropsid-specific CM.  In addition to the hits
identified by the Rfam model we obtained three additional
homologs. Vault RNAs, with a size of about 100 nt, exhibit conserved
sequence patterns only at their ends, with essentially unconstrained
sequence in the central part. Their identification is one of the
well-known and difficult problems for homology search
\cite{Stadler:09b,Kolbe:2009}.

@@

\begin{figure}[ht]
\includegraphics[width=0.475\textwidth]{figures/high-copy-numbers.pdf}
\includegraphics[width=0.475\textwidth]{figures/high-copy-numbers-U6-SRP.pdf}
  \caption[]{{\bf Left:} A comparison of the genomic copy numbers for
    26 of the most abundant RNAs found in human, reptile and the bird
    genomes. {\bf Right:} Bit-score distributions for the Rfam U6,
    Metazoan SRP and the Y RNA covariance models for human and chicken
    paralogues/pseudogenes. }\label{fig:14}
\end{figure}
